The relative roles of genetic, ecological and environmental factors in species radiations

### 2018-02-20 paper/draft.tex
restructure intro

%% Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact that some lineages radiate explosively following
%% island colonisation, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).

%% Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
%% Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs. phylogenetic similarity \cite{losos2008phylogenetic}

@@

%% Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
%% and what trait combinations constitute successful colonizer and radiating species, respectively.

%% [NEW ANGLE:]
% Island biota have always fascinated and inspired researchers (Darwin, Wallace, ...). %TODO: refs
Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics like isolation. (e.g., Losos \& Ricklefs 2009) %TODO: ref
While often featuring unique species communities and harboring countless endemics, %TODO: refs
islands are especially threatened by climate change, human impact and invasive species. %TODO: refs
This makes understanding the functioning and evolution of island communities a high priority objective.

@@

Recently, a group of influential island biogeography researchers proposed a set of 50 questions in island biogeography that still need answering (Patiño et al. 2017).
A subset of these questions deals with the topic of speciation and radiations on islands.
Still, the proposed mechanisms so far include only ecological or geo-morphological processes,
while genetic properties are hardly recognized in the island context.
At the same time, an increasing number of studies investigates genetic processes and
their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein).

@@

In the following, we present a model that remedies both of the adressed shortcomings
by implementing genetic processes on a niche-explicit background
using an island biogeography framework.
We will adress the questions of what relative impact does (a) genetic linkage and
(b) reproductive isolation have on emerging species numbers and diversity,
and (c) which genetic and functional trait characteristics distinguish successful colonizer,
strong competitor and diversifying lineages.

>>>>

@@

%% [NEW ANGLE:]
Islands are understood as suitable model systems for ecology and evolution due to their well definable characteristics like isolation. (e.g., Losos \& Ricklefs 2009) %TODO: ref
While often featuring unique species communities and harboring countless endemics, %TODO: refs
islands are especially threatened by climate change, human impact and invasive species. %TODO: refs
This makes understanding the functioning and evolution of island communities a high priority objective.

@@

Recently, a group of influential island biogeography researchers proposed a set of 50 questions in island biogeography that still need answering (Patiño et al. 2017).
A subset of these questions deals with the topic of speciation and radiations on islands.
%% Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact that some lineages radiate explosively following
%% island colonisation, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).

%% Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
%% Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs. phylogenetic similarity \cite{losos2008phylogenetic}
Still, the proposed mechanisms so far include only ecological or geo-morphological processes,
while genetic properties are hardly recognized in the island context.
At the same time, an increasing number of studies investigates genetic processes and
their effect on reproductive/ecological isolation of populations (e.g., Via \& West 2008, Via 2001 and references therein).

@@

In the following, we present a mechanistic model that remedies both of the adressed shortcomings
by implementing genetic processes on a niche-explicit background
using an island biogeography framework.%% Using an individual-based mechanistic model we investigated how genetic, ecological and environmental factors colonisation and radiation histories on islands,
%% and what trait combinations constitute successful colonizer and radiating species, respectively.
We will adress the questions of what relative impact does (a) genetic linkage and
(b) reproductive isolation have on emerging species numbers and diversity,
and (c) which genetic and functional trait characteristics distinguish successful colonizer,
strong competitor and diversifying lineages.
